Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-29 ms/main.tex
revised corollaries and derivatives on the boundaries; finished summarizing corollaries in discussion; revised spatial model section, figures, and conclusions

\begin{corollary}[Perfect social association of transmission and cooperation]
With perfect social association ($\alpha=1$), the only equilibria are the fixation states, $\tilde{p}=0$ and $\tilde{p}=1$, and cooperation will evolve from any initial frequency (i.e., $\tilde{p}'>\tilde{p}$) if
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
This condition is obtained from inequality \ref{eq:vert_hori_alpha1_condition_proof}, and the right-hand side equals $\gamma_1$ when $\alpha=1$ (\autoref{eq:cost_boundaries}).
Perfect social association (horizontal transmission always occurs during the cooperative interaction) is also assumed in the model studied by \citet{lewin2017microbes}, and therefore this corollary is equivalent to the result in their eq.~1.

\begin{corollary}[No social association of transmission and cooperation]
Without social association ($\alpha=0$), cooperation can take over the population if there is horizontal transmission bias for cooperation, $T_A>T_B$, and if either
\begin{equation}
\label{eq:vert_hori_alpha0}
c < \frac{T_A-T_B}{1-T_B}
\quad \text{or} \quad
\tilde{p}_0 > \tilde{p}^* = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B)} \;.
\end{equation}
\end{corollary}

@@

@@

\begin{corollary}[Condition for cooperation to increase from rarity]
  If the initial frequency of cooperation is very close to zero, $\hat{p}_0 \approx 0$, then its frequency will increase if 
  \begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
  \begin{aligned}
  c < \hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
  \end{aligned}
  \end{equation} 
  \end{corollary}

%Note that the right-hand side equals $\hat\gamma_1$.
This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases maximum cost of cooperation $\hat\gamma_1$ that allows cooperation to increase from rarity (i.e. $\partial \hat\gamma_1 / \partial \alpha > 0$).

@@

In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Then the following applies.
\\

\begin{corollary}[Intermediate association of transmission and cooperation]
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

Interestingly, when $T_A>T_B$ then $\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$, and when $T_A<T_B$ then $\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$.
Therefore, when there is horizontal transmission bias for cooperation, $T_A>T_B$, then vertical transmission hinders the evolution of cooperation ($\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$). In contrast, when there is horizontal transmission bias for defection, vertical transmission facilitates the evolution of cooperation ($\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$).

@@

The order of the interactions across the lattice at each generation is random.
After all interactions take place, and individual's fitness is determined by
$w = 1 + b \cdot n_b - c \cdot n_c$,
where $n_b$ is the number interactions of that individual with cooperative neighbors, and $n_c$ is the total number of interactions that that individual  had ($n_b \le n_c$).
Then a new generation is generated.
Each site is colonized by an offspring of one of the individuals that inhabit either this site or one of the neighboring sites.
The parent is randomly drawn with probability proportional to its fitness, divided by the average fitness in its neighborhood.
Offspring then have the same phenotype as their parents.

\autoref{fig:spatial} shows that the highest cooperation cost to benefit ratio ($c/b$) that permits the evolution of cooperation in simulations of the spatial model is roughly $\alpha \cdot T / (1-T)$, where $T=T_A=T_B$ is the horizontal transmission rate. This agrees with the condition we derived in Corollary~\ref{sec:sym_hor_trans} for a population without population structure, and 
this demonstrates that the social association parameter $\alpha$ can emerge from local interactions between individuals in a structured population.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We studied the evolution of cooperation under non-vertical transmission using deterministic discrete-time evolutionary models with fitnesses in the form of payoffs from a prisoner's dilemma game. 
Under oblique and horizontal cultural transmission, horizontal transmission bias in favor of the cooperative phenotype was found to be necessary and sufficient for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under a combination of vertical, oblique, and horizontal transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the association between cooperation and transmission (Results~\autoref{result:vert_hori} and~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_A<T_B$) and both the cost of cooperation and social association are intermediate (yellow areas in Figures~\ref{fig:result2}-\ref{fig:result3}).

We find that increasing social association ($\alpha$) increases the opportunity for evolution of cooperation (corollary X, \autoref{fig:result2}c and \autoref{fig:result3}cd). 
Similarly, a horizontal transmission bias for cooperation ($T_A-T_B$) will also facilitate its evolution (corollary X, \autoref{fig:result2}).
The effect of vertical/oblique transmission is more complex. 
When there is horizontal transmission bias for cooperation ($T_A>T_B$), increasing the rate of oblique transmission ($1-v$) will facilitate the evolution of cooperation (\autoref{fig:result3}ac). However, when the bias is for defection ($T_A<T_B$), high rates of vertical transmission ($v$) are advantageous for cooperation. 
Nevertheless, to reach fixation of cooperation in the population, there must be a sufficient rate of vertical transmission ($v>\hat v$, Corollary~\autoref{corollary:vhat}).
% TODO carry on summarizing corollaries here

The results that we derived for a deterministic well-mixed model provide a good approximation to results of simulations of a complex stochastic spatial model  (\autoref{fig:spatial}).
In this spatial model, individuals can only interact with and transmit to their neighbors.
This model demonstrates that social association between cooperation and transmission can arise in a structured population in which both types of interactions are local.

@@

%%% Figure: Spatial model
\begin{figure}[h]
  \centering
    \includegraphics[scale=0.7]{spatial.pdf}
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the ratio between the cost and benefit of cooperation ($c/b$) on the y-axis, and the horizontal transmission rate (without transmission bias, $T=T_A=T_B$) on the x-axis.
  The population evolves on a 100-by-100 grid. Selection, cooperation, and horizontal cultural transmission are all local between adjacent sites.
  The black curve represents the condition for the evolution of cooperation in a well-mixed population with social association, $c/b<\alpha T / (1-T)$, where $\alpha=1/8$; see \autoref{eq:equal_transmission}.
  Note that in the structured population, selection is local, whereas in the unstructured population, selection is global. This can explain the small difference in the results.
  Here, population size is $10,000$ (100-by-100 grid); cost of cooperation, $c=0.05$. 50 simulations were executed for each parameter set. Simulations were stopped at generation 10,000 or if one of the phenotypes fixed.
  }
  \label{fig:spatial}
\end{figure}

>>>>

\begin{corollary}[No social association of transmission and cooperation]
\label{corollary:alpha0}
Without social association ($\alpha=0$), cooperation can take over the population if there is horizontal transmission bias for cooperation, $T_A>T_B$, and if either
\begin{equation}
\label{eq:vert_hori_alpha0}
c < \frac{T_A-T_B}{1-T_B}
\quad \text{or} \quad
\tilde{p}_0 > \tilde{p}^* = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B)} \;.
\end{equation}
\end{corollary}

@@

\begin{corollary}[Perfect social association of transmission and cooperation]
\label{corollary:alpha1}
With perfect social association ($\alpha=1$), the only equilibria are the fixation states, $\tilde{p}=0$ and $\tilde{p}=1$, and cooperation will evolve from any initial frequency (i.e., $\tilde{p}'>\tilde{p}$) if
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
This condition is obtained from inequality \ref{eq:vert_hori_alpha1_condition_proof}, and the right-hand side equals $\gamma_1$ when $\alpha=1$ (\autoref{eq:cost_boundaries}).
Perfect social association (horizontal transmission always occurs during the cooperative interaction) is also assumed in the model studied by \citet{lewin2017microbes}, and therefore this corollary is equivalent to the result in their eq.~1.

@@

\begin{corollary}[Condition for cooperation to increase from rarity]
\label{corollary:rarity}
  If the initial frequency of cooperation is very close to zero, $\hat{p}_0 \approx 0$, then its frequency will increase if 
  \begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
  \begin{aligned}
  c < \hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
  \end{aligned}
  \end{equation} 
  \end{corollary}

%Note that the right-hand side equals $\hat\gamma_1$.
This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), making it easier for cooperation to evolve.
However, increasing the horizontal transmission of defection can either increase or decrease the boundary, making it easier or harder for cooperation to evolve, respectively, depending if $T_A$ is high or low (i.e. if $T_A>1/(1+bv\alpha)$ or below).
%($\text{sign}\big(\partial \hat\gamma_1 / \partial T_B \big)= \text{sign}\big(b v \alpha T_A - (1-T_A)\big)$).

@@

In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Note that when $T_A>T_B$ then $\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$, and when $T_A<T_B$ then $\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$.
Similarly, $\partial \hat a_1 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_B > 0$, and $\text{sign}(\partial \hat a_1 / \partial T_B) = \text{sign}\big(T_A (1+b-c) - (1-T_A)cv\big)$.

Then the following applies.
\\

\begin{corollary}[Intermediate association of transmission and cooperation]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

Note that increasing the horizontal transmission of cooperation, $T_A$, decreases both boundaries ($\partial \hat a_1 / \partial T_A, \partial \hat a_2 / \partial T_A < 0$), thus facilitating evolution of cooperation. 
Increasing the horizontal transmission of defection, increases one boundary, ($\partial \hat a_2 / \partial T_B > 0$).
So, if there is horizontal bias for cooperation ($T_A>T_B$), decreasing that bias will make it harder for cooperation to evolve.
However, if there is horizontal bias for defection ($T_A<T_B$) then the effect of further increasing the horizontal transmission of defection, $T_B$, can either increase or decrease the threshold $\hat\alpha_1$, making it harder or easier for cooperation to evolve, respectively, depending if horizontal transmission of cooperation ($T_A$) is high or low (i.e. if $T_A > cv/(1+b-c+cv)$ or below).

Interestingly, when there is horizontal transmission bias for cooperation ($T_A>T_B$), then vertical transmission hinders the evolution of cooperation ($\partial \hat a_2 / \partial v > 0$). In contrast, when there is horizontal transmission bias for defection ($T_A<T_B$), vertical transmission facilitates the evolution of cooperation ($\partial \hat a_1 / \partial v < 0$).
\\

@@

The order of the interactions across the lattice at each generation is random.
After all interactions take place, and individual's fitness is determined by
$w = 1 + b \cdot n_b - c \cdot n_c$,
where $n_b$ is the number interactions of that individual with cooperative neighbors, and $n_c$ is the total number of interactions that that individual  had ($n_b \le n_c$).
Then a new generation is generated, and sites can be settled by offspring of any parent, not just neighbor parents. Thus, selection is global, rather than local, in accordance with our deterministic model.
The parent is randomly drawn with probability proportional to its fitness, divided by the average fitness of all potential parents.
Offspring then have the same phenotype as their parents.

\autoref{fig:spatial} shows that the highest cost of cooperation  ($c$) that permits the evolution of cooperation in simulations of the spatial model roughly agrees with the conditions derived in Result 2. Simulating a spatial model with local selection (i.e. sites can only be settled by offspring of neighbor parents) had only a minor affect, eliminating the stable coexistence.

This comparison between the deterministic well-mixed model and the stochastic spatial model demonstrates that the derived conditions can be useful for predicting the dynamics of complex scenarios. 
Moreover, our spatial model shows how social association ($\alpha$) can emerge from local interactions between individuals in a structured population.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We studied the evolution of cooperation under non-vertical transmission using deterministic discrete-time evolutionary models with fitnesses in the form of payoffs from a prisoner's dilemma game. 
Under oblique and horizontal cultural transmission, horizontal transmission bias in favor of the cooperative phenotype was found to be necessary and sufficient for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under a combination of vertical, oblique, and horizontal transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the association between cooperation and transmission (Results~\autoref{result:vert_hori} and~\autoref{result:vert_obli_hori}).
Importantly, cooperation can increase from rarity (i.e. invade a population of defectors) if and only if (inequality~\ref{eq:unequal_transmission_from_rarity_general_case}),
\begin{equation}
c \cdot v (1-T_B) < b \cdot v \alpha T_A + (T_A - T_B) \;,
\end{equation}
that is, if the effective cost of cooperation (left-hand side) is smaller then the effective benefit plus the horizontal transmission bias (right-hand side).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_A<T_B$) and both the cost of cooperation and social association are intermediate (yellow areas in Figures~\ref{fig:result2}-\ref{fig:result3}).

We find that increasing social association ($\alpha$) increases the opportunity for evolution of cooperation (Corollaries \ref{corollary:intermediate_association} and \ref{corollary:intermediate_association_res3}, Figures~\ref{fig:result2}c and \ref{fig:result3}cd).
Without social association, the benefit of cooperation cannot facilitate its evolution; cooperation can only succeed under horizontal transmission bias ($T_A>T_B$, Corollary~\ref{corollary:alpha0}). 
Indeed, horizontal transmission plays a major role in the evolution of cooperation. Mostly, increasing the transmission of cooperation ($T_A$) or decreasing the transmission of defection ($T_B$) facilitates the evolution of cooperation (Corollaries~\ref{corollary:rarity} and~\ref{corollary:intermediate_association_res3}, \autoref{fig:result2}). 
However, in specific cases increasing the transmission of defection can be advantageous for cooperation (Corollaries~\ref{corollary:rarity} and~\ref{corollary:intermediate_association_res3}).
The effect of oblique transmission is more complex (Corollary~\ref{corollary:intermediate_association_res3}). 
When there is horizontal transmission bias for cooperation ($T_A>T_B$), increasing the rate of oblique transmission ($1-v$) will facilitate the evolution of cooperation (\autoref{fig:result3}ac).
In contrast, when the bias is for defection ($T_A<T_B$), high rates of vertical transmission ($v$) are advantageous for cooperation, and there must be sufficient rate of vertical transmission ($v>\hat v$, Corollary~\autoref{corollary:vhat}, \autoref{fig:result3}bd) for cooperation to fix in the population.

The conditions derived from our deterministic well-mixed model provide a good approximation to results of simulations of a complex stochastic spatial model  (\autoref{fig:spatial}).
In this spatial model, individuals can only interact with and transmit to their neighbors.
This model demonstrates that social association between cooperation and transmission can arise in a structured population in which both types of interactions are local.

@@

%%% Figure: Spatial model
\begin{figure}[h]
  \centering
    \includegraphics[scale=0.5]{spatial1.png}
    ~
    \includegraphics[scale=0.5]{spatial2.png} \\
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of the left panel, or the transmission bias $T_A-T_B$ on the x-axis of the right panel.
  The population evolves on a 100-by-100 grid. Cooperation and horizontal cultural transmission are all local between adjacent sites, and each site had 8 neighbors.
  The black curves represent the cost boundaries for the evolution of cooperation in a well-mixed population with social association where $\alpha=1/8$ in (left) \autoref{eq:equal_transmission} and (right) \label{eq:cost_boundaries}.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed and 50 simulations were executed for each parameter set.
  Here, population size is $10,000$ (100-by-100 grid), benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$.
  \textbf{(left)} Symmetric horizontal transmission, $T=T_A=T_B$. \textbf{(right)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$.
  }
  \label{fig:spatial}
\end{figure}
